---
title: "User Manual: powerHLA"
author: "SK"
date: "Oct 20, 2021"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{User Manual: powerHLA}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r}
require(powerHLA)
require(rstan)
require(knitr)
require(parallel)
require(ggplot2)
require(ggforce)
require(plot.matrix)
```

# Introduction

The 4 Mb major histocompatibility complex (MHC) region on chromosome 6,
which encodes the human leukocyte antigen (HLA) genes, is highly
polymorphic, and specific HLA genes have been implicated in increased
risk for autoimmune and infectious diseases.

Humans inherit two sets of parental HLA genes (alleles) from each HLA
class (e.g. A, B, etc.). Hence, in a population of size $N$ there are
$2N$ alleles from each HLA class. To determine whether a specific HLA
allele is associated with increased risk/protection with respect to a
certain disease, we compare the frequencies (prevalence, $\theta$) of
that allele between a cohort of disease-positive individuals and a
matching cohort of normal individuals. For quantitative identification
of HLA alleles with enriched/depleted frequencies we need **statistical
modeling**.

`r BiocStyle::Githubpkg(repo = "snaketron/powerHLA", pkg = "powerHLA")`
helps the user address some of the following aspects of HLA association
analyses:

-   quantification of allele frequencies in a given population
-   simulation of HLA allele counts data
-   power analysis

# Methods

## Simulation of HLA allele frequencies

In a population with $N$ individuals we have $2N$ alleles from a
specific HLA class (e.g. HLA class A). The HLA alleles are distributed
across $K$ HLA types (categories), i.e. we observe a vector of $K$
counts $\vec{n}=n_1,n_2,...,n_K$, with $n_i \in \mathbb{Z}^{\geq0}$. HLA
allele counts data is **compositional** as the HLA allele count $n_i$ is
part of a whole $\sum_i^K n_i = 2N$.

We can simulate HLA allele counts using a **multinomial distribution**
with parameters $\vec{\theta}=\theta_1, \theta_2, ..., \theta_K$
distributed on a K-simplex, with $\theta_i \in \mathbb{R}^{>0}$ and
$\sum^{K}_i\theta_i=1$. To simulate HLA allele counts from a Multinomial
distribution we have to provide two pieces of information:

-   $\vec{\theta}$: allele frequencies
-   $N$: population size = the number of alleles in the simulated
    population

For this task we will use HLA class A allele frequencies from a
population of 3,456,066 unrelated donors with self-reported German
descent who have been recruited by the German Bone Marrow Donor File
(DKMS). This data was downloaded from The Allele Frequency Net Database
(<http://www.allelefrequencies.net>).

The data is provided as part of
`r BiocStyle::Githubpkg(repo = "snaketron/powerHLA", pkg = "powerHLA")`.
Inspect it now using the next few lines of code:

```{r}
# get from hla allele freq net
data("dkms_de")
knitr::kable(head(dkms_de))
```

Column `af` of dkms_de provides the allele frequencies ($\theta$) and
column `n` the number of individuals in the dataset.

```{r}
# number of HLA alleles in population
N <- dkms_de$n[1]*2

# number of distinct HLA alleles (categories)
K <- nrow(dkms_de)

# allele frequencies
theta <- dkms_de$af
```

The $\theta$s are reported as 4-digit fractions. Hence, rounding errors
are possible. Check whether $\sum_i^K \theta_i = 1$.

```{r}
sum(theta)
```

To account for the missing $\theta=0.0056$, such that $\vec{\theta}$
satisfies the criterion $\sum_i^K \theta_i = 1$, we add one additional
HLA allele type called `other` with $\theta_{\text{other}}=0.0056$

```{r}
other <- data.frame(allele = "other",
                    pop = dkms_de$pop[1],
                    af = 1-sum(dkms_de$af),
                    n = dkms_de$n[1])

dkms_de <- rbind(dkms_de, other)

sum(dkms_de$af)
```

Finally, we will update the variables K, N and theta

```{r}
# number of distinct HLA alleles (categories)
K <- nrow(dkms_de)

# number of HLA alleles in population
N <- dkms_de$n[1]*2

# allele frequencies
theta <- dkms_de$af
```

We can also plot $\vec{\theta}$

```{r, fig.width=4.5, fig.height=3}
plot(theta, ylab = expression(theta), xlab = "HLA allele")
```

Additionally, we can estimate the allele counts, i.e. how many times
each allele occurs in this datasets: $\gamma = \theta \cdot N$.

Due to the rounding error that was mentioned earlier, $\gamma$ contains
fractions instead of integers $\rightarrow$ let's round them up to the
nearest one.

```{r}
gamma <- round(x = theta*N, digits = 0)

# we are off by 5 alleles, but we can live with 
# this given that N > 6*10^6 
sum(gamma)-N
```

Now lets simulate $B=10$ samples of HLA allele counts based on the observed 
$\vec{\theta}$. The simulated populations will have size similar to the size 
of the original population $N=7*10^6$ alleles ($3.5*10^6$ individuals).

```{r}
rng_gamma <- powerHLA::get_multinomial_rng(theta = theta, B = 10, N = 6912132)
```

The variable `rng_gamma` contains $B=10$ rows (one for each sample) and $K$ 
columns (one for each allele). The cells are simulated allele counts
given $\vec{\theta}$ and $N$.

```{r}
dim(rng_gamma)
```

As anticipated, we see an overall agreement between the $\vec{\gamma}$ from 
`rng_gamma` and the observed (background) $\vec{\gamma}$

```{r, fig.width=6, fig.height=6}
par(mfrow=c(2,2))
plot(rng_gamma[1, ], gamma);abline(a = 0, b = 1);
plot(rng_gamma[2, ], gamma);abline(a = 0, b = 1);
plot(rng_gamma[3, ], gamma);abline(a = 0, b = 1);
plot(rng_gamma[4, ], gamma);abline(a = 0, b = 1);
```

We can also visualize the matrix `rng_gamma` as a tile plot:

```{r, fig.width=7, fig.height=5}
par(mar=c(5.1, 4.1, 4.1, 4.1))
plot(log10(rng_gamma))
```

## Quantification of allele frequencies in a sample

```{r}

```



## Power analysis

Power is the probability of detecting an effect ($\delta$), given that the 
effect is really there. Performing power analysis is essential to determine 
the necessary number of subjects needed to detect an effect of a given size.

To perform power analysis with powerHLA the user has to provide the following
pieces of information:

  * effect size $\delta$: e.g. $\delta=0.1$ (1% change in allele frequency)
  * background allele frequencies: $\vec{\theta}$
  * background allele counts: $\vec{\gamma}$
  * number of simulations to determine power: $B$
  * number of posterior draws: $P$
  * Dirichlet prior: $\vec{\alpha}$
  * population size(s): $N$ (number of alleles in simulated population)

```{r}

p <- powerHLA::get_power_analysis(
  ns = c(100, 500, 1000),
  theta = theta,
  gamma = gamma,
  B = 500,
  rng_draws = 2000,
  a = rep(x = 0.1, times = K),
  cores = 3,
  verbose = T)

```

```{r, fig.width=8, fig.height=6}

ggplot(data = p)+
  facet_wrap(facets = ~N, ncol = 2)+
  geom_errorbar(aes(x = allele, y = diff_mean, ymin = diff_X2.5.,
                    ymax = diff_X97.5., group = B),
             position = position_dodge(width = 0.3),
             col = "gray", width = 0.05)+
  geom_point(aes(x = allele, y = diff_mean, group = B),
             position = position_dodge(width = 0.3), size = 0.25)+
  scale_x_discrete(limits = paste0("K", 1:K), 
                   breaks = paste0("K", seq(from = 1, to = K, by = 4)))+
  theme_bw(base_size = 10)+
  ylab(expression(delta))

```

```{r}

p$diff_effect_HDI95 <- ifelse(test = p$diff_X2.5.<=0 & p$diff_X97.5.>=0, yes = 0, no = 1)
p$diff_min_effect_95 <- apply(X = p[,c("diff_X2.5.", "diff_X97.5.")], MARGIN = 1,
                              FUN = function(x) {return(min(abs(x)))})*p$diff_effect_HDI95
q <- aggregate(diff_effect_HDI95~N+allele, data = p, FUN = sum)
q$diff_effect_HDI95_pct <- q$diff_effect_HDI95/max(p$B)*100

```

```{r, fig.width=6, fig.height=3}

ggplot(data = q)+
  geom_point(aes(y = diff_effect_HDI95_pct, x = N, col = allele))+
  geom_line(aes(y = diff_effect_HDI95_pct, x = N, col = allele))+
  ylim(c(0, 100))+
  theme_bw()+
  ylab(label = "% effect found")

```

```{r, fig.width=7, fig.height=6}

ggplot(data = p)+
  facet_wrap(facets = ~N)+
  geom_sina(aes(x = allele, y = diff_min_effect_95), size = 0.25)+
  scale_x_discrete(limits = paste0("K", 1:K))+
  theme_bw(base_size = 10)+
  ylab(expression("|"*delta*"|"[min]))+
  scale_y_log10()

```

# Session

```{r}
sessionInfo()
```
